---
title: "Experiment 3 - Energy Intake"
author: "Brent Bachman"
date: 2025-05-01

editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib

format: gfm
toc: true
number-sections: true
fig-format: png
fig-width: 6
fig-asp: 0.618
fig-align: center
fig-dpi: 300
---

# Packages

```{r}
#| label: load-packages
#| output: false

library("tidyverse")
```

```{r}
#| label: install-packages
#| eval: false

# install.packages("tidyverse")
```

```{r}
#| label: session-info
sessionInfo()
```

# Data

## Import, tidy, and transform

```{r}
#| label: mydata

# Create a dataframe
mydata <- 
  
  # Import the data
  read_csv(
    file ="data/experiment-3-statsdata.csv",
    show_col_types = FALSE
    ) |>
  
  # Select the variables of interest 
  # EI = energy intake
  select(id, age, cohort, sex, diet, EI_0:EI_53) |>
  
  # Divide cohort B's energy intake values on day 34 by 2 (since they were recorded over two days) 
  # mutate(
  #   EI_34 = if_else(cohort == 1, EI_34/2, EI_34)
  #   ) |>
  
  # Tidy data
  pivot_longer(
    cols = starts_with("EI"),
    names_to = "day",
    values_to = "energy_intake",
    values_drop_na = TRUE
    ) |>
  
  # Transform data
  mutate(
  
    # Recode categorical variables as factors
    id = factor(id),
    cohort = factor(
      cohort, 
      levels = c(0, 1),
      labels = c("Cohort A", "Cohort B")
    ),
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Male", "Female")
      ),
    diet = factor(
      diet,
      levels = c(0, 1, 2),
      labels = c("CHOW", "HFD-PALM", "HFD-COCO")
      ),
    
    # Recode sex and cohort using sum coding
    sex_sum    = C(sex, sum),
    cohort_sum = C(cohort, sum),
    
    # Compute time in weeks
    day = parse_number(day),
    week = day/7,
    .before = energy_intake
    ) |>
  
  # Drop na values
  drop_na()

# View a summary of the data
summary(mydata)
```

```{r}
# Compute the mean and SD of each cohort's age
mydata |>
  group_by(cohort) |>
  summarize(
    age_m  = mean(age),
    age_sd = sd(age)
  )
```

## Visualize - Line Plot

```{r}
#| label: line-plot

# Create a line plot of each group's daily energy intake over time
line_plot <-
  mydata |>  
  ggplot(
    aes(
      x = week, 
      y = energy_intake, 
      color = diet, 
      shape = diet
      )
    ) +
  
  # Facet by sex and cohort
  facet_grid(
    sex ~ cohort,
    scales = "free_y"
    ) + 
  
  # Plot the observed means as points
  stat_summary(
    fun = mean,
    geom = "point"
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar"
    ) +
  
  # Plot connecting lines
  stat_summary(
    fun = mean,
    geom = "line"
    ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange3",
      "HFD-COCO"  = "darkorchid3"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change y-axis title
  labs(y = "Energy Intake (kcal)") +
  
  # Change x-axis title and scale
  scale_x_continuous(
    name = "Time (Weeks)",
    breaks = seq(from = 0, to = 8, by = 1)) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Rotate and adjust the x-axis tick labels
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 0.5,
      hjust = 0.5
      ),
    legend.position = "top"
    )

# Show the plot
line_plot
```

In both cohorts, the data for CHOW seems pretty flat over time. However, the data for HFD-PALM and, to a lesser degree, HFD-COCO, seem to spike on day 1 then slowly decrease over time. (There is also a peak for all groups around week 5 in cohort B; this is because data was not collected the day before, so those energy intake values are computed over two days).

# Summarize

```{r}
#| label: model-data

# Create a new dataframe
model_data <- 
  
  # Copy the original dataframe
  mydata |>
  
  # Keep just one observation per subject
  distinct(
    id, 
    .keep_all = TRUE
  ) |>
  
  # Deselect variables of disinterest
  select(-day, -week, -energy_intake) |>
  
  # Join that dataframe to a new dataframe
  left_join(
    
    # Copy the original dataframe
    mydata |>
  
    # filter the dataframe to include only the dietary intervention
    filter(day > 0) |>

    # Compute each subject's average energy intake across the dietary intervention
    group_by(id) |>
    summarize(
      average_energy_intake = mean(energy_intake)
    ) |>
    ungroup(),
  
  # join by the key of subject id
  by = join_by(id)
  )

# Show a summary of the data
summary(model_data)
```

## Visualize - Bar Plot

Let's visualize each group's average energy intake across days 1 to 53.

```{r}
#| label: bar-plot

# Create a bar plot each group's average energy intake across the entire dietary intervention
bar_plot <-
  model_data |>  
  ggplot(
    aes(
      x     = diet, 
      y     = average_energy_intake, 
      color = diet, 
      fill  = diet,
      shape = diet
      )
    ) +
  
  # Facet by cohort and sex
  facet_grid(
    sex ~ cohort,
    # scales = "free_y"
    ) + 
  
  # Plot the observed means as points
  stat_summary(
    fun = mean,
    geom = "bar",
    color = "black",
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    color = "black",
    width = 0.2
    ) +
  
  # Plot individual data points
  geom_point(
    position = position_jitter(
      width = 0.2
    )
  ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange4",
      "HFD-COCO"  = "darkorchid4"
      )
    ) +
  
  # Change fill title and scale
  scale_fill_manual(
    name = "Diet",
    values = c(
      "CHOW"     = "ivory3",
      "HFD-PALM" = "darkorange2",
      "HFD-COCO" = "darkorchid2"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change x-axis title
  labs(x = "Diet") +
  
  # Change y-axis title and scale
  scale_y_continuous(
    name = "Average Energy Intake (kcal/day)",
    limits = c(0, 150),
    breaks = seq(from = 0, to =  150, by = 25)
  ) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Remove x-axis title, text, and tick labels
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
    )

# Show the plot
bar_plot
```

Consistent with the earlier visualization, average energy intake seems to be greater for HFD-PALM and, perhaps to a *slightly* lesser degree, HFD-COCO, in both sexes in both cohorts.

# Model

To test the effects of sex, diet, and their interaction on average energy intake across the dietary intervention, a multiple linear regression model will be built with the predictors sex (sum-coded: male = 1 and female = -1), diet (treatment-coded: CHOW = 0; HFD-PALM and HFD-COCO = 1), cohort (treatment-coded: Cohort A = 0 and Cohort B = 1), as well as the two- and three-way interactions between each predictor.

```{r}
#| label: model

# Build the model
model <- 
  lm(
    average_energy_intake ~ cohort_sum * sex_sum * diet, 
    data = model_data
  )
```

## Assumptions

Before I run any statistical tests, let's check how well the model satisfies the assumptions.

First, I need to add the fitted and residual values to the data.

```{r}
#| label: model-fits

# Add the fitted and residual values to the dataset
model_fits <- 
  model_data |>
  mutate(
    fits   = c(fitted(model)),
    resids = c(residuals(model))
  )
```

### Linearity and Homoskedasticity

Next, I will create a residuals plot to check for linearity and homoskedasticity.

```{r}
#| label: residuals-plot

# Create a residuals plot to check for linearity and homoskedasticity
residuals_plot <-
  model_fits |>
  ggplot(
    aes(x = fits, y = resids)
  ) +
  
  # Plot individual data points
  geom_point() +
  
  # Plot a horizontal line at y = 0
  geom_hline(yintercept = 0, col = "red") + 
  
  # Change aesthetics
  labs(
    title = "Residuals vs Predicted Values",
    x = "Predicted Values",
    y = "Residuals"
  ) +
  theme_bw() 

# Show the plot
residuals_plot
```

The data seem to be evenly distributed along the horizontal axis, indicating that the data are approximately linear; and the data seem to deviate along the y-axes (i.e., from the horizontal red line) evenly across the x-axis, indicating that the data are homoskedastic.

### Normality

Next, I will create a QQ plot to check for normality.

```{r}
#| label: qq-plot

# Create a QQ plot to check for normality
qq_plot <-
  model_fits |>
  ggplot(
    aes(sample = resids)
  ) +
  
  # Plot boxplots
  geom_qq() +
  geom_qq_line(col = "red") +
  
  # Change aesthestics
  labs(
    title = "Normal QQ Plot",
    x = "Theoretical",
    y = "Sample"
  ) +
  theme_bw() 

# Show the plot
qq_plot
```

The data seem to fall mostly along the red line, indicating that the data are approximately normally distributed.

## Omnibus Tests

With the model built, let's perform omnibus tests for the effects of each predictor using the function "anova()". This function uses type I sum of squares. Thus, it will produce F test statistics for the effects of each predictor entered sequentially (i.e., the residual effect of each predictor after accounting for the effects of all the other predictors entered in the model before it).

```{r}
# Perform omnibus tests
model |> anova()
```

When predicting average energy intake over the course of the dietary intervention, there were significant effects of cohort, $F(1, 62) = 24.42, p < .001$, sex, $F(1, 62) = 222.04, p < .001$, and diet, $F(2, 62) = 19.83, p < .001$. However, after controlling for these predictors, there were not significant effects of the two-way interactions between cohort and sex, $F(1, 62) = 0.00, p = .980$, cohort and diet, $F(2, 62) = 0.63, p = .539$, or sex and diet, $F(2, 62) = 2.08, p = .134$, or of the three-way interaction between cohort, sex, and diet, $F(2, 62) = 1.42, p = .249$.

## Summary, Referece = CHOW

Let's probe these effects by producing the summary output.

```{r}
# Produce summary output
model |> summary()
```

Overall, the model explained 83% of the variance in average energy intake; and, after adjusting for the number of predictors, it explained 80%, $F(11, 62) = 26.76, R^2 = 0.83, R^2_{adj} = 0.80, p < .001$.

Within the male CHOW subjects, cohort B consumed significantly less energy on average per day than cohort A ($B = -12.59, SE = 3.37, p < .001$), probably because they were slightly older ($M_{A} = 110, SD_{A} = 0.58, M_{B} = 152, SD_{B} = 1.14$). However, the difference between the two cohort's CHOW groups was not significantly different between the sexes ($B = 0.54, SE = 3.37, p = .874$); across the sexes, the difference between the two cohorts was not significantly different in either HFD-PALM ($B = 3.05, SE = 4.68, p = .517$) or HFD-COCO ($B = 5.31, SE = 4.77, p = .270$) relative to CHOW; and the magnitude of the cohort differences (or lack thereof) for both HFD-PALM ($B = 2.76, SE = 4.68, p = .558$) and HFD-COCO ($B = 5.06, SE = 4.77, p = .293$) relative to CHOW was not significantly different between the sexes. These results indicate that the cohorts consumed significantly different amounts of energy on average per day across the entire dietary intervention, but the difference is consistent across sex and diet groups.

Across cohorts and across the sexes, both HFD-PALM ($B = 14.26, SE = 2.34, p < .001$) and HFD-COCO ($B = 10.60, SE = 2.38, p < .001$) consumed significantly more energy on average per day than CHOW. The magnitude of the difference between HFD-PALM and CHOW did not significantly interact with cohort ($B = -1.52, SE = 2.34, p = .517$), sex ($B = 4.35, SE = 2.34, p = .068$), or cohort and sex ($B = -1.38, SE = 2.34, p = .558$). The magnitude of the difference between HFD-COCO and CHOW also did not significantly interact with cohort ($B = -2.65, SE = 2.38, p = .270$), sex ($B = 3.76, SE = 2.38, p = .120$), or cohort and sex ($B = 2.53, SE = 2.38, p = .293$).

## Summary, Reference = HFD-PALM

To compare HFD-COCO with HFD-PALM, let's relevel the diet variable to make HFD-PALM the reference group, then refit the model and produce the summary output. (There is no need to rerun the omnibus tests, as the "total" effects are already captured, and thus the output would be identical).

```{r}
#| label: model-palm

# Relevel factors to change reference group to HFD-VS
data_palm <- 
  model_data |>
  mutate(
    diet = fct_relevel(diet, "HFD-PALM"), # relevel diet
  )

# Build the model
model_palm <- 
  model |>
  update(
    data = data_palm
  )

# Produce the summary output
model_palm |> summary()
```

Across cohorts and across the sexes, although HFD-COCO consumed less energy on average per day than HFD-PALM, this difference was not statistically significant ($B = -3.66, SE = 2.34, p = .123$); and the difference between HFD-COCO and HFD-PALM (or lack thereof) did not significantly interact with cohort ($B = -1.13, SE = 2.34, p = .631$), sex ($B = -0.59, SE = 2.34, p = .803$), or sex and cohort ($B = 3.91, SE = 2.34, p = .100$).

and although the magnitude of this difference (or lack thereof) was smaller in males than females, this difference was also not statistically significant ($B = 3.32, SE = 3.37, p = .329$).

# Communicate

Let's save the plots.

```{r}
#| label: save-plots
#| eval: false

# Save the ...

# line plot
ggsave(
  plot = line_plot,
  filename = "output/experiment-3/01-energy-intake/01-line-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# bar plot
ggsave(
  plot = bar_plot,
  filename = "output/experiment-3/01-energy-intake/02-bar-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# residuals plot
ggsave(
  plot = residuals_plot,
  filename = "output/experiment-3/01-energy-intake/03-residuals-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# qq plot
ggsave(
  plot = qq_plot,
  filename = "output/experiment-3/01-energy-intake/04-qq-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)
```

```{r}
#| label: save-rds-file
#| eval: false

# Save the bar plot as an rds file to be imported later and arranged into a single multi-panel plot with all the body composition data
saveRDS(
  bar_plot,
  file = "output/experiment-3/05-body-comp-figure/01-energy-intake.rds"
)
```

# References
