---
title: "Experiment 1 - Fat Mass"
author: "Brent Bachman"
date: 2025-05-01

editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib

format: gfm
toc: true
number-sections: true
fig-format: png
fig-width: 6
fig-asp: 0.618
fig-align: center
fig-dpi: 300
---

# Packages

```{r}
#| label: load-packages
#| output: false

library("tidyverse")
```

```{r}
#| label: install-packages
#| eval: false

# install.packages("tidyverse")
```

```{r}
#| label: session-info
sessionInfo()
```

# Data

## Import, tidy, and transform

```{r}
#| label: mydata

# Create a dataframe called mydata
mydata <- 
  
  # Import the data
  read_csv(
    file = "data/experiment-1-statsdata.csv",
    show_col_types = FALSE
    ) |>
  
  # Select the variables of interest 
  # ei = energy intake
  select(id, cohort, sex, diet, fm_0:fm_21) |>
  
  # Exclude subject 7
  filter(id != 7) |>
  
  # Tidy data
  pivot_longer(
    cols = starts_with("fm"),
    names_to = "day",
    values_to = "fat_mass",
    values_drop_na = TRUE
    ) |>
  
  # Transform data
  mutate(
  
    # Recode categorical variables as factors
    id = factor(id),
    cohort = factor(
      cohort,
      levels = c(0, 1),
      labels = c("Ad Libitum", "Pair-Fed")
      ),
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Male", "Female")
      ),
    diet = factor(
      diet,
      levels = c(0, 1, 2, 3),
      labels = c("CHOW", "HFD-VS", "HFD-COCO", "HFD-C8/10")
      ),
    
    # Recode sex using sum coding
    sex_sum = C(sex, sum),
    
    # Compute time in weeks
    day = parse_number(day),
    week = day/7,
    .before = fat_mass,
    )

# View a summary of the data
summary(mydata)
```

## Line Plot

Let's visualize the data.

```{r}
#| label: line-plot

# Create a line plot of each subject's fat mass over time
line_plot <-
  mydata |>  
  ggplot(
    aes(
      x = week, 
      y = fat_mass, 
      color = diet, 
      shape = diet
      )
    ) +
  
  # Facet by sex and cohort
  facet_grid(
    sex ~ cohort,
    scales = "free_y"
    ) + 
  
  # Plot the observed means as points
  stat_summary(
    fun = mean,
    geom = "point"
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.2
    ) +
  
  # Plot connecting lines
  stat_summary(
    fun = mean,
    geom = "line"
    ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-VS"    = "firebrick3",
      "HFD-COCO"  = "darkorchid3",
      "HFD-C8/10" = "dodgerblue3"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-VS"    = 16,
      "HFD-COCO"  = 17,
      "HFD-C8/10" = 18
      )
    ) +
  
  # Change y-axis title and scale
  labs(y = "Fat Mass (g)") +
  
  # Change x-axis title and scale
  scale_x_continuous(
    name = "Time (Weeks)",
    breaks = seq(from = 0, to = 4, by = 1)) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Rotate and adjust the x-axis tick labels
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 0.5,
      hjust = 0.5
      )
    )

# Show the plot
line_plot
```

The data seem pretty flat over time for chow and HFD-C8/10 in both males and females in the ad libitum cohort, as well as in all diet groups in both sexes in the pair-fed cohort. However, the slope seems to be greater in male and female HFD-VS and HFD-COCO in the ad libitum cohort.

# Summarize

Let's calculate the slope of each subject's fat mass over time.

```{r}
#| label: model-data

# Create a new data frame called model_data
model_data <-
  
  # Copy the original dataframe
  mydata |>
  
  # Group by subject id
  group_by(id) |>
  
  # Filter the data to include only days 0 or 21
  filter(day == 0 | day == 21) |>

  # Compute the slope of body weight
  mutate(
    fat_mass_slope = 
      (fat_mass - lag(fat_mass)) / (day - lag(day))
  ) |>
  
  # Ungroup by id
  ungroup() |> 
  
  # Select only the relevant columns
  select(
    id, cohort, sex, sex_sum, diet, fat_mass_slope
  ) |>
  
  # Drop na values
  drop_na(fat_mass_slope)

# Show a summary of the data
summary(model_data)
```

Looks good.

## Bar Plot

Now, let's visualize the slopes.

```{r}
#| label: bar-plot

# Create a bar plot of each subject's fat mass slope
bar_plot <-
  model_data |>  
  ggplot(
    aes(
      x     = diet, 
      y     = fat_mass_slope, 
      color = diet, 
      fill  = diet,
      shape = diet
      )
    ) +
  
  # Facet by sex and cohort
  facet_grid(
    sex ~ cohort,
    ) + 
  
  # Plot the observed means as points
  stat_summary(
    fun = mean,
    geom = "bar",
    color = "black",
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    color = "black",
    width = 0.2
    ) +
  
  # Plot individual data points
  geom_point(
    position = position_jitter(
      width = 0.2
    )
  ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-VS"    = "firebrick4",
      "HFD-COCO"  = "darkorchid4",
      "HFD-C8/10" = "dodgerblue4"
      )
    ) +
  
  # Change fill title and scale
  scale_fill_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "ivory3",
      "HFD-VS"    = "firebrick2",
      "HFD-COCO"  = "darkorchid2",
      "HFD-C8/10" = "dodgerblue2"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-VS"    = 16,
      "HFD-COCO"  = 17,
      "HFD-C8/10" = 18
      )
    ) +
  
  # Change x-axis title
  labs(x = "Diet") +
  
  # Change y-axis title and scale
  scale_y_continuous(
    name = "Fat Mass Slope (g/day)",
    limits = c(-2, 6),
    breaks = seq(from = -2, to = 6, by = 2)
  ) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Remove x-axis tick labels
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
    )

# Show the plot
bar_plot
```

Consistent with the previous visualization, it looks like HFD-VS has a greater fat mass slope than all other groups in the ad libitum cohort, but not the pair-fed cohort. It looks like HFD-COCO has a greater fat mass slope than CHOW and HFD-C8/10 in the ad libitum cohort, but not the pair-fed cohort; and it looks like the fat mass slope of HFD-C8/10 is not different from CHOW in either cohort.

# Model

To test the effects of sex, diet, cohort, and their interactions on the slope of fat mass over time, a multiple linear regression will be built to predict fat mass gain (in g/day) with sex (sum-coded, Levels: male = 1, female = -1), diet (treatment coded, levels: CHOW, HFD-VS, HFD-COCO, HFD-C8/10), cohort (treatment coded, levels: ad libitum and pair-fed) and their interactions.

```{r}
#| label: model

# Build a linear model with the outcome variable average energy intake and the predictors sex, diet, cohort, and the interaction between diet and cohort
model <- lm(fat_mass_slope ~ sex_sum + diet * cohort, data = model_data)
```

## Assumptions

Before I run any statistical tests, let's check how well the model satisfies the assumptions. 

But first, I need to add the fitted and residual values to the data.

```{r}
#| label: model-data-fits

# Add the fitted and residual values to the dataset
model_data_fits <- 
  model_data |>
  mutate(
    fits = c(fitted(model)),
    resids = c(residuals(model))
  )
```

### Linearity and Homoskedasticity

Next, let's create a residuals vs fitted plot to check for linearity and homoskedasticity.

```{r}
#| label: residuals-plot

# Create a residuals plot to check for linearity and homoskedasticity
residuals_plot <-
  model_data_fits |>
  ggplot(
    aes(x = fits, y = resids)
  ) +
  
  # Plot individual data points
  geom_point() +
  
  # Plot a horizontal line at y = 0
  geom_hline(yintercept = 0, col = "red") + 
  
  # Change aesthetics
  labs(
    title = "Residuals vs Predicted Values",
    x = "Predicted Values",
    y = "Residuals"
  ) +
  theme_bw() 

# Show the plot
residuals_plot
```

The data seem approximately linear and homoskedastic.

### Normality

Finally, let's create a qq plot to check for normality.

```{r}
#| label: qq-plot

# Create a QQ plot to check for normality
qq_plot <-
  model_data_fits |>
  ggplot(
    aes(sample = resids)
  ) +
  
  # Plot boxplots
  geom_qq() +
  geom_qq_line(col = "red") +
  
  # Change aesthestics
  labs(
    title = "Normal QQ Plot",
    x = "Theoretical",
    y = "Sample"
  ) +
  theme_bw() 

# Show the plot
qq_plot
```

The data seem approximately normally distributed.

## Omnibus Tests

Now that we know this is a good model, let's perform omnibus tests for the effects of each predictor using the function "anova()". This function uses type I sum of squares. Thus, it will produce F test statistics for the effects of each predictor entered sequentially (i.e., the residual effect of each predictor after accounting for the effects of all the other predictors entered in the model before it).

```{r}
# Perform omnibus tests
model |> anova()
```

When predicting body weight gain over the course of the dietary intervention, there were significant effects of sex, $F(1, 52) = 11.90, p = .001$, diet, $F(3, 52) = 21.51, p < .001$, cohort, $F(1, 52) = 46.96, p < .001$, and the interaction between diet and cohort, $F(1, 52) = 12.92, p < .001$. 

## Reference = CHOW

Let's probe these effects by producing the summary output.

```{r}
# Produce summary output
model |> summary()
```

Overall, the model explained 76% of the variance in body weight gain; and, after adjusting for the number of predictors, it explained 72%, $F(8, 52) = 20.27, R^2 = 0.76, R^2_{adj} = 0.72, p < .001$.

Across sex, the difference between the ad libitum and pair-fed cohorts' CHOW groups was not statistically significant ($B = 0.01, SE = 0.32, p = .963$). However, both HFD-VS ($B = 2.52, SE = 0.30, p < .001$) and HFD-COCO ($B = 1.64, SE = 0.32, p < .001$) gained significantly more fat mass per day than CHOW in the ad libitum cohort. The magnitude of the difference between CHOW and both HFD-VS ($B = -2.43, SE = 0.44, p < .001$) and HFD-COCO ($B = -1.55, SE = 0.45, p = .001$) was significantly reduced in the pair-fed cohort. Although HFD-C8/10 gained slightly *less* fat mass per day than CHOW in the ad libitum cohort, this effect was not statistically significant ($B = -0.02, SE = 0.30, p = .948$); and it was not significantly different in the pair-fed cohort ($B = 0.38, SE = 0.44, p = .395$).

## Reference = HFD-VS

To directly compare HFD-COCO and HFD-C8/C10 to HFD-VS, let's relevel the diet variable to make HFD-VS the reference group, then refit the model and produce the summary output. (There is no need to rerun the omnibus tests, as the "total" effects are already captured, and thus the output would be identical).

```{r}
#| label: model-vs

# Relevel factors to change reference group to HFD-VS
model_data_vs <- 
  model_data |>
  mutate(
    diet = fct_relevel(diet, "HFD-VS"), # relevel diet
  )

# Build the model
model_vs <- 
  model |>
  update(
    data = model_data_vs
  )

# Produce the summary output
model_vs |> summary()
```

Both HFD-COCO ($B = -0.87, SE = 0.32, p = .008$) and HFD-C8/10 ($B = -2.54, SE = 0.30, p < .001$) gained significantly less fat mass per day than HFD-VS in the ad libitum cohort. The magnitude of the difference between HFD-VS and HFD-C8/10 ($B = 2.05, SE = 0.43, p < .001$) but not HFD-COCO ($B = 0.87, SE = 0.45, p = .055$) was significantly reduced in the pair-fed cohort.

## Reference = HFD-COCO

To directly compare HFD-C8/C10 to HFD-COCO, let's relevel the diet variable to make HFD-COCO the reference group, then refit the model and produce the summary output. (There is no need to rerun the omnibus tests, as the "total" effects are already captured, and thus the output would be identical).

```{r}
#| label: model-coco

# Relevel factors to change reference group to HFD-VS
model_data_coco <- 
  model_data |>
  mutate(
    diet = fct_relevel(diet, "HFD-COCO"), # relevel diet
  )

# Build the model
model_coco <- 
  model |>
  update(
    data = model_data_coco
  )

# Produce the summary output
model_coco |> summary()
```

HFD-C8/10 gained significantly less fat mass per day than HFD-COCO in the ad libitum cohort ($B = -1.66, SE = 0.32, p < .001$); and the magnitude of this difference was reduced in the pair-fed cohort ($B =  1.18, SE = 0.45, p = .011$).

# Communicate

Let's save the plots.

```{r}
#| label: save-plots
#| eval: false

# Save the ...

# line plot
ggsave(
  plot = line_plot,
  filename = "output/experiment-1/03-fat-mass/01-line-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# bar plot
ggsave(
  plot = bar_plot,
  filename = "output/experiment-1/03-fat-mass/02-bar-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# residuals plot
ggsave(
  plot = residuals_plot,
  filename = "output/experiment-1/03-fat-mass/03-residuals-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# qq plot
ggsave(
  plot = qq_plot,
  filename = "output/experiment-1/03-fat-mass/04-qq-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)
```

```{r}
#| label: save-rds-file
#| eval: false

# Save the bar plot as an rds file to be imported later and arranged into a single multi-panel plot with all the body composition data
saveRDS(
  bar_plot,
  file = "output/experiment-1/05-body-comp-figure/03-fat-mass.rds"
)
```

# References
