---
title: "Experiment 3 - iba1"
author: "Brent Bachman"
date: 2025-05-02

editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib

format: gfm
toc: true
number-sections: true
fig-format: png
fig-width: 6
fig-asp: 0.618
fig-align: center
fig-dpi: 300
---

# Packages

```{r}
#| label: load-packages
#| output: false

library("tidyverse")
library("ggpubr")
```

```{r}
#| label: install-packages
#| eval: false

# install.packages("tidyverse")
# install.packages("ggpubr")
```

```{r}
#| label: session-info
sessionInfo()
```

# Data

## Import, tidy, and transform

```{r}
#| label: mydata

# Create a dataframe called mydata
mydata <- 
  
  # Import the data
  read_csv(
    file ="data/experiment-3-statsdata-iba1.csv",
    show_col_types = FALSE
    ) |>
  
  # Select the variables of interest 
  select(id, sex, diet, cohort, 
         arc_iba1_number_avg, arc_iba1_density_avg,
         ca1_iba1_number_avg, ca1_iba1_density_avg
         ) |>
  
  # Transform data
  mutate(
  
    # Recode categorical variables as factors
    id = factor(id),
    cohort = factor(
      cohort,
      levels = c(0, 1),
      labels = c("A", "B")
      ),
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Male", "Female")
      ),
    diet = factor(
      diet,
      levels = c(0, 1, 2),
      labels = c("CHOW", "HFD-PALM", "HFD-COCO")
      ),
    
    # Recode sex using sum coding
    sex_sum = C(sex, sum)
    ) |>
  
  drop_na()

# View a summary of the data
summary(mydata)
```

## Sample Size

Since some of the brain data was lost/unusable, let's compute the grouped sample sizes of the usable dataset

```{r}
#| label: sample-size-total

# Compute total sample size
mydata |> 
  summarize(
    count = n()
  )
```

```{r}
#| label: sample-size-grouped

# Compute grouped sample sizes
mydata |>
  group_by(sex, diet) |>
  summarize(
    count = n()
  )
```

Looks like the total sample size is 36 and the grouped sample sizes are 6 per group.

## Visualize

### ARC Number

```{r}
#| label: arc-number-plot

# Create a plot object
arc_number_plot <-
  
  # Call the dataset
  mydata |>  
  
  # Plot each group's iba1+ cell number in the ARC
  ggplot(
    aes(
      x = diet, 
      y = arc_iba1_number_avg, 
      fill = diet,
      )
    ) +
  
  # Create a plot title
  ggtitle(label = "ARC") +
  
  # Facet by sex (rows)
  facet_grid(
    . ~ sex
    ) + 
  
  # Plot the observed means as bars
  stat_summary(
    fun.data = mean_se,
    geom = "bar",
    color = "black",
    position = position_dodge(0.95)
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    color = "black",
    width = 0.2,
    position = position_dodge(0.95)
    ) +
  
  # Plot the individual data points
  geom_point(
    aes(
      color = diet, 
      shape = diet
      ),
    position = position_jitterdodge(
      jitter.width = 0.4,
      jitter.height = 0,
      dodge.width = 0.95
      )
    ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange4",
      "HFD-COCO"  = "darkorchid4"
      )
    ) +
  
  # Change fill title and scale
  scale_fill_manual(
    name = "Diet",
    values = c(
      "CHOW"     = "ivory3",
      "HFD-PALM" = "darkorange2",
      "HFD-COCO" = "darkorchid2"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change y-axis title and scale
  scale_y_continuous(
    name = "Iba1+ cells per section",
    limits = c(0, 700),
    breaks = seq(from = 0, to = 700, by = 100)
  ) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Rotate and adjust the x-axis tick labels
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
    )

# Show the plot
arc_number_plot
```

### ARC Density

```{r}
#| label: arc-density-plot

# Create a plot object
arc_density_plot <-
  
  # Call the dataset
  mydata |>  
  
  # Plot each group's iba1 cell density in the ARC
  ggplot(
    aes(
      x = diet, 
      y = arc_iba1_density_avg, 
      fill = diet,
      )
    ) +
  
  # Create a plot title
  ggtitle(label = "MBH") +
  
  # Facet by sex (rows)
  facet_grid(
    . ~ sex
    ) + 
  
  # Plot the observed means as bars
  stat_summary(
    fun.data = mean_se,
    geom = "bar",
    color = "black",
    position = position_dodge(0.95)
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    color = "black",
    width = 0.2,
    position = position_dodge(0.95)
    ) +
  
  # Plot the individual data points
  geom_point(
    aes(
      color = diet, 
      shape = diet
      ),
    position = position_jitterdodge(
      jitter.width = 0.4,
      jitter.height = 0,
      dodge.width = 0.95
      )
    ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange4",
      "HFD-COCO"  = "darkorchid4"
      )
    ) +
  
  # Change fill title and scale
  scale_fill_manual(
    name = "Diet",
    values = c(
      "CHOW"     = "ivory3",
      "HFD-PALM" = "darkorange2",
      "HFD-COCO" = "darkorchid2"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change y-axis title and scale
  scale_y_continuous(
    name = bquote("Iba1+ cells per" ~ mm^3),
    limits = c(0, 24000),
    breaks = seq(from = 0, to = 24000, by = 4000)
  ) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Rotate and adjust the x-axis tick labels
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
    )

# Show the plot
arc_density_plot
```

### CA1 Number

```{r}
#| label: ca1-number-plot

# Create a plot object
ca1_number_plot <-
  
  # Call the dataset
  mydata |>  
  
  # Plot each group's iba1 cell number in the ca1
  ggplot(
    aes(
      x = diet, 
      y = ca1_iba1_number_avg, 
      fill = diet,
      )
    ) +
  
  # Create a plot title
  ggtitle(label = "CA1") +
  
  # Facet by sex (rows)
  facet_grid(
    . ~ sex
    ) + 
  
  # Plot the observed means as bars
  stat_summary(
    fun.data = mean_se,
    geom = "bar",
    color = "black",
    position = position_dodge(0.95)
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    color = "black",
    width = 0.2,
    position = position_dodge(0.95)
    ) +
  
  # Plot the individual data points
  geom_point(
    aes(
      color = diet, 
      shape = diet
      ),
    position = position_jitterdodge(
      jitter.width = 0.4,
      jitter.height = 0,
      dodge.width = 0.95
      )
    ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange4",
      "HFD-COCO"  = "darkorchid4"
      )
    ) +
  
  # Change fill title and scale
  scale_fill_manual(
    name = "Diet",
    values = c(
      "CHOW"     = "ivory3",
      "HFD-PALM" = "darkorange2",
      "HFD-COCO" = "darkorchid2"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change y-axis title and scale
  scale_y_continuous(
    name = "Iba1+ cells per section",
    limits = c(0, 700),
    breaks = seq(from = 0, to = 700, by = 100)
  ) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Rotate and adjust the x-axis tick labels
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
    )

# Show the plot
ca1_number_plot
```

### CA1 Density

```{r}
#| label: ca1-density-plot

# Create a plot object
ca1_density_plot <-
  
  # Call the dataset
  mydata |>  
  
  # Plot each group's iba1+ cell density in the CA1 region
  ggplot(
    aes(
      x = diet, 
      y = ca1_iba1_density_avg, 
      fill = diet,
      )
    ) +
  
  # Create a plot title
  ggtitle(label = "CA1") +
  
  # Facet by sex (rows)
  facet_grid(
    . ~ sex
    ) + 
  
  # Plot the observed means as bars
  stat_summary(
    fun.data = mean_se,
    geom = "bar",
    color = "black",
    position = position_dodge(0.95)
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    color = "black",
    width = 0.2,
    position = position_dodge(0.95)
    ) +
  
  # Plot the individual data points
  geom_point(
    aes(
      color = diet, 
      shape = diet
      ),
    position = position_jitterdodge(
      jitter.width = 0.4,
      jitter.height = 0,
      dodge.width = 0.95
      )
    ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange4",
      "HFD-COCO"  = "darkorchid4"
      )
    ) +
  
  # Change fill title and scale
  scale_fill_manual(
    name = "Diet",
    values = c(
      "CHOW"     = "ivory3",
      "HFD-PALM" = "darkorange2",
      "HFD-COCO" = "darkorchid2"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change y-axis title and scale
  scale_y_continuous(
    name = bquote("Iba1+ cells per" ~ mm^3),
    limits = c(0, 15000),
    breaks = seq(from = 0, to = 15000, by = 3000)
  ) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Rotate and adjust the x-axis tick labels
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
    )

# Show the plot
ca1_density_plot
```

# Model 1: ARC

To test the effects of sex, diet, and their interaction on iba1+ cell density in the ARC, a multiple linear regression model will be built with the predictors sex (sum-coded: male = 1 and female = -1), diet (treatment coded: CHOW = 0; HFD-PALM and HFD-COCO = 1), and the interaction between sex and diet.

```{r}
#| label: model-arc

# Build the model
model_arc <- lm(
  arc_iba1_density_avg ~ sex_sum * diet,
  data = mydata
)
```

## Assumptions

Before I run any statistical tests, let's check how well the model satisfies the assumptions.

First, I need to add the fitted and residual values to the data.

```{r}
#| label: model-arc-fits

# Add the fitted and residual values to the dataset
model_arc_fits <- 
  mydata |>
  mutate(
    fits = c(fitted(model_arc)),
    resids = c(residuals(model_arc))
  )
```

### Linearity and Homoskedasticity

Next, I will create a residuals plot to check for linearity and homoskedasticity.

```{r}
#| label: model-arc-residuals-plot

# Create a residuals plot to check for linearity and homoskedasticity
model_arc_residuals_plot <-
  model_arc_fits |>
  ggplot(
    aes(x = fits, y = resids)
  ) +
  
  # Plot individual data points
  geom_point() +
  
  # Plot a horizontal line at y = 0
  geom_hline(yintercept = 0, col = "red") + 
  
  # Change aesthetics
  labs(
    title = "Residuals vs Predicted Values",
    x = "Predicted Values",
    y = "Residuals"
  ) +
  theme_bw() 

# Show the plot
model_arc_residuals_plot
```

The data seem to be linear and homoskedastic.

### Normality

Next, I will create a QQ plot to check for normality.

```{r}
#| label: model1-qq-plot

# Create a QQ plot to check for normality
model_arc_qq_plot <-
  model_arc_fits |>
  ggplot(
    aes(sample = resids)
  ) +
  
  # Plot points and qq line
  geom_qq() +
  geom_qq_line(col = "red") +
  
  # Change aesthestics
  labs(
    title = "Normal QQ Plot",
    x = "Theoretical",
    y = "Sample"
  ) +
  theme_bw() 

# Show the plot
model_arc_qq_plot
```

The data seem to be approximately normally distributed.

## Omnibus Tests

With the model built, let's perform omnibus tests for the effects of each predictor using the function "anova()". This function uses type I sum of squares. Thus, it will produce F test statistics for the effects of each predictor entered sequentially (i.e., the residual effect of each predictor after accounting for the effects of all the other predictors entered in the model before it).

```{r}
#| label: model-arc-omnibus

# Perform omnibus F tests
model_arc |> anova()
```

There were not significant effects of sex, $F(1, 30) = 0.16, p = .694$, diet, $F(2, 30) = 1.04, p = .367$, or an interaction between sex and diet, $F(2, 30) = 0.81, p = .455$.

## Summary, Reference = CHOW

Let's take a look at the summary output.

```{r}
# Produce summary output
model_arc |> summary()
```

The model did not explain a significant portion of the variance in iba1+ cell density in the ARC, $F(5, 30) = 0.77, R^2 = 0.11, R^2_{adj} = -0.04, p = .579$.

# Model 2: CA1

To test the effects of sex, diet, and their interaction on iba1+ cell density in the CA1 region, a multiple linear regression model will be built with the predictors sex (sum-coded: male = 1 and female = -1), diet (treatment coded: CHOW = 0; HFD-PALM and HFD-COCO = 1), and the interaction between sex and diet.

```{r}
#| label: model-ca1

# Build the model
model_ca1 <- lm(
  ca1_iba1_density_avg ~ sex_sum * diet,
  data = mydata
)
```

## Assumptions

Before I run any statistical tests, let's check how well the model satisfies the assumptions.

First, I need to add the fitted and residual values to the data.

```{r}
#| label: model-ca1-fits

# Add the fitted and residual values to the dataset
model_ca1_fits <- 
  mydata |>
  mutate(
    fits = c(fitted(model_ca1)),
    resids = c(residuals(model_ca1))
  )
```

### Linearity and Homoskedasticity

Next, I will create a residuals plot to check for linearity and homoskedasticity.

```{r}
#| label: model-ca1-residuals-plot

# Create a residuals plot to check for linearity and homoskedasticity
model_ca1_residuals_plot <-
  model_ca1_fits |>
  ggplot(
    aes(x = fits, y = resids)
  ) +
  
  # Plot individual data points
  geom_point() +
  
  # Plot a horizontal line at y = 0
  geom_hline(yintercept = 0, col = "red") + 
  
  # Change aesthetics
  labs(
    title = "Residuals vs Predicted Values",
    x = "Predicted Values",
    y = "Residuals"
  ) +
  theme_bw() 

# Show the plot
model_ca1_residuals_plot
```

The data seem to be linear and homoskedastic.

### Normality

Next, I will create a QQ plot to check for normality.

```{r}
#| label: model-ca1-qq-plot

# Create a QQ plot to check for normality
model_ca1_qq_plot <-
  model_ca1_fits |>
  ggplot(
    aes(sample = resids)
  ) +
  
  # Plot points and qq line
  geom_qq() +
  geom_qq_line(col = "red") +
  
  # Change aesthestics
  labs(
    title = "Normal QQ Plot",
    x = "Theoretical",
    y = "Sample"
  ) +
  theme_bw() 

# Show the plot
model_ca1_qq_plot
```

The data seem to be approximately normally distributed.

## Omnibus Tests

With the model built, let's perform omnibus tests.

```{r}
# Perform omnibus F tests
model_ca1 |> anova()
```

There were not significant effects of sex, $F(1, 30) = 0.04, p = .852$, diet, $F(2, 30) = 0.22, p = .802$, or an interaction between sex and diet, $F(2, 30) = 0.56, p = .579$.

## Summary, Reference = CHOW

Let's take a look at the summary output.

```{r}
# Produce summary output
model_ca1 |> summary()
```

The model did not explain a significant portion of the variance in iba1+ cell density in the CA1 region, $F(5, 30) = 0.32, R^2 = 0.05, R^2_{adj} = -0.11, p = .898$.

# Communicate

Let's combine the plots

```{r}
#| label: arranged-plot

# Combine the plots for each brain region into a single, multi-panel plot
arranged_plot <- 
  ggarrange(
    arc_density_plot, ca1_density_plot,
    nrow = 1, ncol = 2,
    # labels = c("A", "B", "C", "D"),
    common.legend = T
  )

# Show the plot
arranged_plot
```

Now let's save the plots.

```{r}
#| label: save-plots
#| eval: false

# Save the ...

# MBH density plot
ggsave(
  plot = arc_density_plot,
  filename = "output/experiment-3/07-iba1/01-MBH-density-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# MBH residuals plot
ggsave(
  plot = model_arc_residuals_plot,
  filename = "output/experiment-3/07-iba1/02-MBH-residuals-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# MBH qq plot
ggsave(
  plot = model_arc_qq_plot,
  filename = "output/experiment-3/07-iba1/03-MBH-qq-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# CA1 plot
ggsave(
  plot = ca1_density_plot,
  filename = "output/experiment-3/07-iba1/04-CA1-density-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# CA1 residuals plot
ggsave(
  plot = model_ca1_residuals_plot,
  filename = "output/experiment-3/07-iba1/05-CA1-residuals-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# CA1 qq plot
ggsave(
  plot = model_ca1_qq_plot,
  filename = "output/experiment-3/07-iba1/06-CA1-qq-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# arranged plot - 300 dpi
ggsave(
  plot = arranged_plot,
  filename = "output/experiment-3/07-iba1/07-arranged-plot-300.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# arranged plot - 1200 dpi
ggsave(
  plot = arranged_plot,
  filename = "output/experiment-3/07-iba1/08-arranged-plot-1200.tiff",
  width = 6, height = 3.708, units = "in", dpi = 1200
)
```