---
title: "Experiment 3 - Fat Mass"
author: "Brent Bachman"
date: 2025-05-01

editor: visual
editor_options: 
  chunk_output_type: console
bibliography: references.bib

format: gfm
toc: true
number-sections: true
fig-format: png
fig-width: 6
fig-asp: 0.618
fig-align: center
fig-dpi: 300
---

# Packages

```{r}
#| label: load-packages
#| output: false

library("tidyverse")
```

```{r}
#| label: install-packages
#| eval: false

# install.packages("tidyverse")
```

```{r}
#| label: session-info
sessionInfo()
```

# Data

## Import, tidy, and transform

```{r}
#| label: mydata

# Create a dataframe
mydata <- 
  
  # Import the data
  read_csv(
    file ="data/experiment-3-statsdata.csv",
    show_col_types = FALSE
    ) |>
  
  # Select the variables of interest 
  # BW = body weigh
  select(id, cohort, sex, diet, FM_m7, FM_53) |>
  
  # Tidy data
  pivot_longer(
    cols = starts_with("FM"),
    names_to = "day",
    values_to = "fat_mass",
    values_drop_na = TRUE
    ) |>
  
  # Transform data
  mutate(
  
    # Recode categorical variables as factors
    id = factor(id),
    cohort = factor(
      cohort, 
      levels = c(0, 1),
      labels = c("Cohort A", "Cohort B")
    ),
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Male", "Female")
      ),
    diet = factor(
      diet,
      levels = c(0, 1, 2),
      labels = c("CHOW", "HFD-PALM", "HFD-COCO")
      ),
    
    # Recode sex using sum coding
    sex_sum = C(sex, sum),
    
    # Compute time in weeks
    day = parse_number(day),
    day = if_else(
      day == 7,
      -7, 
      53
      ),
    week = day/7,
    .before = fat_mass
    ) |>
  
  # Drop na values
  drop_na()

# View a summary of the data
summary(mydata)
```

## Visualize - line plot

```{r}
#| label: line-plot

# Create a line plot of each group's fat mass over time
line_plot <-
  
  # Call the dataset
  mydata |>  
  
  # Initialize plot parameters
  ggplot(
    aes(
      x = week, 
      y = fat_mass, 
      color = diet, 
      shape = diet
      )
    ) +
  
  # Facet by sex and cohort
  facet_grid(
    sex ~ cohort,
    scales = "free_y"
    ) + 
  
  # Plot the observed means as points
  stat_summary(
    fun = mean,
    geom = "point"
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.2
    ) +
  
  # Plot connecting lines
  stat_summary(
    fun = mean,
    geom = "line"
    ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange3",
      "HFD-COCO"  = "darkorchid3"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change y-axis title and scale
  labs(y = "Fat Mass (g)") +
  
  # Change x-axis title and scale
  scale_x_continuous(
    name = "Time (Weeks)",
    breaks = seq(from = -1, to = 8, by = 1)) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Rotate and adjust the x-axis tick labels
  theme(
    axis.text.x = element_text(
      angle = 45,
      vjust = 0.5,
      hjust = 0.5
      ),
    legend.position = "top"
    )

# Show the plot
line_plot
```

In both cohorts, the data seem to increase over time across groups. However, it seems to increase at a faster rate for both HFD groups.

# Summary

First, let's calculate each subject's fat mass slope across the dietary intervention.

```{r}
#| label: model-data

# Create a new data frame
model_data <-
  
  # Copy the original dataframe
  mydata |>

  # Compute each subject's fat mass slope
  group_by(id) |>
  mutate(
    fat_slope = 
      (fat_mass - lag(fat_mass)) / (day - lag(day))
  ) |>
  ungroup() |> 
  
  # Select only the relevant columns
  select(
    id, cohort, sex, sex_sum, diet, fat_slope
  ) |>
  
  # Drop na values
  drop_na(fat_slope)

# Show a summary of the data
summary(model_data)
```

## Visualize - bar plot

Let's visualize the model data.

```{r}
#| label: bar-plot

# Create a plot object
bar_plot <-
  
  # Call the dataset
  model_data |>  
  
  # Plot each group's average energy intake
  ggplot(
    aes(
      x     = diet, 
      y     = fat_slope, 
      color = diet, 
      fill  = diet,
      shape = diet
      )
    ) +
  
  # Facet by sex (rows)
  facet_grid(
    cohort ~ sex
    ) + 
  
  # Plot the observed means as points
  stat_summary(
    fun = mean,
    geom = "bar",
    color = "black",
    ) +
  
  # Plot the observed standard errors as errorbars
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    color = "black",
    width = 0.2
    ) +
  
  # Plot individual data points
  geom_point(
    position = position_jitter(
      width = 0.2
    )
  ) +
  
  # Change color title and scale
  scale_color_manual(
    name = "Diet",
    values = c(
      "CHOW"      = "black",
      "HFD-PALM"  = "darkorange4",
      "HFD-COCO"  = "darkorchid4"
      )
    ) +
  
  # Change fill title and scale
  scale_fill_manual(
    name = "Diet",
    values = c(
      "CHOW"     = "ivory3",
      "HFD-PALM" = "darkorange2",
      "HFD-COCO" = "darkorchid2"
      )
    ) +
  
  # Change shape title and scale
  scale_shape_manual(
    name = "Diet",
    values = c(
      "CHOW"      = 15,
      "HFD-PALM"  = 16,
      "HFD-COCO"  = 17
      )
    ) +
  
  # Change x-axis title
  labs(x = "Diet") +
  
  # Change y-axis title and scale
  scale_y_continuous(
    name = "Fat Mass Slope (g/day)",
    limits = c(-1, 4), 
    breaks = seq(from = -1, to =  4, by = 1)
  ) +
  
  # Change overall plot theme
  theme_bw() + 
  
  # Remove x-axis title, text, and tick labels
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
    )

# Show the plot
bar_plot
```

Consistent with the earlier visualization, fat mass slope seems to be greater for HFD-PALM and, perhaps to a *slightly* lesser degree, HFD-COCO, in both sexes in both cohorts.

# Model

To test the effects of sex, diet, and their interaction on the linear slope of fat mass across the dietary intervention, a multiple linear regression model will be built with the predictors sex (sum-coded: male = 1 and female = -1), diet (treatment-coded: CHOW = 0; HFD-PALM and HFD-COCO = 1), and the interaction between sex and diet.

```{r}
#| label: model

# Build the model
model <- lm(
  fat_slope ~ sex_sum * diet, 
  data = model_data
  )
```

## Assumptions

Before I run any statistical tests, let's check how well the model satisfies the assumptions.

First, I need to add the fitted and residual values to the data.

```{r}
#| label: model-fits

# Add the fitted and residual values to the dataset
model_fits <- 
  model_data |>
  mutate(
    fits = c(fitted(model)),
    resids = c(residuals(model))
  )
```

### Linearity and Homoskedasticity

Next, I will create a residuals plot to check for linearity and homoskedasticity.

```{r}
#| label: residuals-plot

# Create a residuals plot to check for linearity and homoskedasticity
residuals_plot <-
  model_fits |>
  ggplot(
    aes(x = fits, y = resids)
  ) +
  
  # Plot individual data points
  geom_point() +
  
  # Plot a horizontal line at y = 0
  geom_hline(yintercept = 0, col = "red") + 
  
  # Change aesthetics
  labs(
    title = "Residuals vs Predicted Values",
    x = "Predicted Values",
    y = "Residuals"
  ) +
  theme_bw() 

# Show the plot
residuals_plot
```

The data seem to be evenly distributed along the horizontal axis, indicating that the data are approximately linear; and the data seem to deviate along the y-axes (i.e., from the horizontal red line) evenly across the x-axis, indicating that the data are homoskedastic.

### Normality

Next, I will create a QQ plot to check for normality.

```{r}
#| label: qq-plot

# Create a QQ plot to check for normality
qq_plot <-
  model_fits |>
  ggplot(
    aes(sample = resids)
  ) +
  
  # Plot boxplots
  geom_qq() +
  geom_qq_line(col = "red") +
  
  # Change aesthestics
  labs(
    title = "Normal QQ Plot",
    x = "Theoretical",
    y = "Sample"
  ) +
  theme_bw() 

# Show the plot
qq_plot
```

The data seem to fall mostly along the red line, indicating that the data are approximately normally distributed.

## Omnibus Tests

With the model built, let's perform omnibus tests for the effects of each predictor using the function "anova()". This function uses type I sum of squares. Thus, it will produce F test statistics for the effects of each predictor entered sequentially (i.e., the residual effect of each predictor after accounting for the effects of all the other predictors entered in the model before it).

```{r}
# Perform omnibus tests
model |> anova()
```

When predicting fat mass slope across the dietary intervention, there were significant effects of sex, $F(1, 32) = 47.59, p < .001$, and diet, $F(2, 32) = 44.95, p < .001$, but not an interaction between sex and diet, $F(2, 32) = 2.30, p = .116$.

## Summary, Reference = CHOW

Let's probe these effects by producing the summary output.

```{r}
# Produce summary output
model |> summary()
```

Overall, the model explained 82% of the variance in fat mass slope across the dietary intervention; and, after adjusting for the number of predictors, it explained 79%, $F(5, 32) = 28.42, R^2 = 0.82, R^2_{adj} = 0.79, p < .001$.

Across the sexes, both HFD-PALM ($B = 1.04, SE = 0.12, p < .001$) and HFD-COCO ($B = 0.95, SE = 0.12, p  <.001$) gained significantly more fat mass per day than CHOW. The magnitude of the difference between CHOW and HFD-PALM ($B = 0.25, SE = 0.12, p = .045$) but not HFD-COCO ($B = 0.08, SE = 0.12, p = .506$) was significantly greater in males than females.

## Summary, Reference = HFD-PALM

To compare HFD-COCO with HFD-PALM, let's relevel the diet variable to make HFD-PALM the reference group, then refit the model and produce the summary output. (There is no need to rerun the omnibus tests, as the "total" effects are already captured, and thus the output would be identical).

```{r}
#| label: model-palm

# Relevel factors to change reference group to HFD-VS
data_palm <- 
  model_data |>
  mutate(
    diet = fct_relevel(diet, "HFD-PALM"), # relevel diet
  )

# Build the model
model_palm <- 
  model |>
  update(
    data = data_palm
  )

# Produce the summary output
model_palm |> summary()
```

Across the sexes, although HFD-COCO gained less fat mass per day than HFD-PALM, this difference was not statistically significant ($B = -0.09, SE = 0.12, p = .457$); and although the magnitude of this difference was greater in males than females, this difference was also not statistically significant ($B = -0.17, SE = 0.12, p = .173$).

# Communicate

Let's save the plots.

```{r}
#| label: save-plots
#| eval: false

# Save the ...

# line plot
ggsave(
  plot = line_plot,
  filename = "output/experiment-3/03-fat-mass/01-line-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# bar plot
ggsave(
  plot = bar_plot,
  filename = "output/experiment-3/03-fat-mass/02-bar-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# residuals plot
ggsave(
  plot = residuals_plot,
  filename = "output/experiment-3/03-fat-mass/03-residuals-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)

# qq plot
ggsave(
  plot = qq_plot,
  filename = "output/experiment-3/03-fat-mass/04-qq-plot.tiff",
  width = 6, height = 3.708, units = "in", dpi = 300
)
```

```{r}
#| label: save-rds-file
#| eval: false

# Save the bar plot as an rds file to be imported later and arranged into a single multi-panel plot with all the body composition data
saveRDS(
  bar_plot,
  file = "output/experiment-3/05-body-comp-figure/03-fat-mass.rds"
)
```

# References
